Very simplified likelihood

### 2024-02-28 inst/very_simplified_likelihood.Rmd
in silico formula 1

## Placing last

@@

## Note on the differences between probability and likelihood

@@

# Formula 1.

@@

## Mathematical Olympiad

@@

## Parkrun

>>>>

# Placing last

@@

# Note on the differences between probability and likelihood

@@

# Formula 1: Ocon vs Gasly

@@

# Formula 1: Formula 1 drivers: career arc

```{r getnumgetfoc}
readstring <- function(year){read.table(paste("formula1_",year,".txt",sep=""))}
getfoc <- function(year,comp="Perez"){  # get focal competitor
  M <- as.matrix(readstring(year))
  out <- suppressWarnings(as.numeric(M[comp,seq_len(ncol(M)-1)]))
  out[is.na(out)] <- nrow(M)
  return(out)
}
getnum <- function(year){nrow(readstring(year))} # number of competitors

perez <- lapply(2011:2023,getfoc)
print(perez)
y <- unlist(lapply(seq_along(perez),function(i){mean(perez[[i]])}))
x <- 2011:2023
plot(x,y)
summary(lm(y~x))
```

```{r checofirst}
checo_like <- function(a){
  out <- a*0
  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year)-1,getnum(year)-1,log=TRUE) }
  return(out)
}

a <- seq(from=0.45,to=0.62,by=0.01)
cL <- checo_like(a)
cL <- cL - max(cL)
plot(a,cL,type='b')
abline(h=c(0,-2))
```

```{r definef1log}
f1_logistic <- function(vec){
  alpha <- vec[1]
  beta  <- vec[2]
  out <- 0
  for(year in 2011:2023){
     LO <- alpha + beta*(year-2011)
     strength <- exp(LO)/(1+exp(LO))
     out <- out + f_vec_arn(strength,getfoc(year)-1,getnum(year)-1,log=TRUE)
  }
  return(out)
}
```

```{r calculatelogisticcontou,cache=TRUE}
a <- seq(from=-0.8,to=0.0,by=0.01)
b <- seq(from=0,to=0.2,by=0.01)
jj <- as.matrix(expand.grid(a,b))
L <- apply(jj,1,f1_logistic)
L <- L-max(L)
L <- matrix(L,length(a),length(b))
```

```{r showfilledc}
L <- pmax(L,-40)
contour(a,b,L)
```

```{r uselogf1}
f1_logistic(c(0,0))
f1_logistic(c(0,0.0001))
f1_logistic(c(0.001,0))
```

```{r optimizelof1,cache=TRUE}
o <- optim(c(-0.2,0.1),fn=f1_logistic, control=list(fnscale = -1),hessian=TRUE)
```

```{r showjj}
o
o$par
o$hessian
eigen(o$hessian)
best <- o$par
f1_logistic(best)

jj <- best
jj[1] <- jj[1]*1.01
f1_logistic(jj) - f1_logistic(best)

jj <- best
jj[1] <- jj[1]*0.99
f1_logistic(jj) - f1_logistic(best)

jj <- best
jj[2] <- jj[2]*1.00001
f1_logistic(jj) - f1_logistic(best)

jj <- best
jj[2] <- jj[2]*0.99
f1_logistic(jj) - f1_logistic(best)
```

```{r testlogf1,cache=TRUE}
o_free <- optim(c(-0.2,0.1),fn=f1_logistic, control=list(fnscale = -1),hessian=TRUE)
o_constrained <- optim(c(-0.2,0.1),fn=function(v){
v[2] <- 0
return(f1_logistic(v))}, control=list(fnscale = -1),hessian=TRUE)
```

```{r showpvalue}
pchisq(2*(o_free$value - o_constrained$value),df=1,lower.tail=FALSE)
```

# in silico formula

```{r insn}
table(replicate(100,which(rrace3(pn=c(a=1,b=10),ps=c(a=0.9,b=0.1))=='a')))
```

```{r insilicoestmation,cache=TRUE}
n <- 1000
M <- matrix(0,n,4)
alpha_true <- 4
beta_true <- -6
for(i in seq_len(n)){
    x <- i/n
    LO <- alpha_true + beta_true*x
    p <- exp(LO)/(1+exp(LO))
    nn <- round(runif(1,10,100))
    M[i,1] <- x  # regressor
    M[i,2] <- which(rrace3(pn=c(a=1,b=nn),ps=c(a=p,b=1-p))=='a')
    M[i,3] <- nn+1
    M[i,4] <- p
}
head(M)
tail(M)
```

```{r testran,cache=FALSE}
runif(10)
```

```{r definelikeinsilico}
likeinsilico <- function(vec,M){
  alpha <- vec[1]
  beta  <- vec[2]
  S <- 0
  for(i in seq_len(nrow(M))){
    x <- M[i,1]
    LO <- alpha + beta * x
    p <- exp(LO)/(1+exp(LO))
    S <- S + f_vec_arn(p,M[i,2]-1,M[i,3]-1,log=TRUE)
  }
return(S)
}
```

```{r testlikeininsilicco,cache=TRUE}
likeinsilico(c(4,-6),M)
likeinsilico(c(4,-6.5),M)
likeinsilico(c(4,-5.5),M)
```

```{r optimize_insilicco,cache=TRUE}
optim(par=c(4,-6),likeinsilico,control=list(fnscale= -1),M=M)
```

```{r makeZcontourdataset,cache=TRUE}
a <- seq(from=3.8,to=4.2,by=0.1)
b <- seq(from=-6.1,to=-5.6,by=0.1)
Z <- as.matrix(expand.grid(a,b))
Z <- apply(Z,1,function(x){likeinsilico(x,M=M)})
Z <- matrix(Z,length(a),length(b))
```

```{r doZcontour}
Z <- Z-max(Z)
contour(a,b,Z,levels=-c(0:10,15,20,30))
points(alpha_true,beta_true,pch=16,cex=4)
grid()
persp(a,b,Z)
```

# Mathematical Olympiad

@@

# Parkrun
